cardinalR: Generating interesting high-dimensional data structures

### 2025-05-06 paper-cardinalR.Rmd
Added initial pyr descriptions

@@

@@

@@

@@

>>>>

The `gen_pyrrect(n, p, h, l_vec, rt)` function generates a dataset of $n$ points forming a high-dimensional pyramid-like structure with a rectangular base. The pyramid is embedded in a $p$-dimensional space, with a tip at height zero and base at height $h$. The shape tapers linearly from the base dimensions $(l\_x, l\_y)$ to a smaller rectangular section with side lengths $(2r\_t, 2r\_t)$ at the tip. Points are distributed more densely near the tip, simulating a natural skew toward smaller height values.

Let $x\_1, x\_2, \dots, x\_p$ denote the coordinates of the generated points. The final dimension $x\_p$ encodes the height of each point and is drawn from an exponential distribution capped at $h$:

$$
x_p = z \sim \min\left(\text{Exp}(\lambda = 2/h),\ h\right)
$$.

Let $r\_x(z)$ and $r\_y(z)$ denote the half-widths of the rectangular cross-section at height $z$:

$$
r_x(z) = r_t + (l_x - r_t) \cdot \frac{z}{h}, \quad
r_y(z) = r_t + (l_y - r_t) \cdot \frac{z}{h}
$$.

The first three coordinates are then defined as:

$$
x_1 \sim \mathcal{U}(-r_x(z),\ r_x(z)), \quad
x_2 \sim \mathcal{U}(-r_y(z),\ r_y(z)), \quad
x_3 \sim \mathcal{U}(-r_x(z),\ r_x(z))
$$.

For $p > 3$, the remaining $p - 4$ dimensions (i.e., $x\_4$ to $x\_{p-1}$) are tapered toward zero with decreasing height and generated as:

$$
x_j \sim \mathcal{U}(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right), \quad \text{for } j = 4, \dots, p-1.
$$

This tapering ensures that the structure narrows in all directions as it approaches the tip.

@@

The `gen_pyrtri(n, p, h, l, rt)` function generates a dataset of $n$ points forming a high-dimensional pyramid-like structure with a triangular cross-section. The structure is embedded in a $p$-dimensional space, with the tip located at height 0 and the base at height $h$. The triangle expands linearly in size from tip to base, with more points concentrated near the tip.

Let $x\_1, x\_2, \dots, x\_p$ denote the coordinates of the generated points. The final coordinate \$x\_p\$ encodes the height $z$ of each point and is drawn from an exponential distribution capped at $h$:

$$
x_p = z \sim \min\left(\text{Exp}(\lambda = 2/h),\ h\right).
$$

Let $r(z)$ denote the scaling factor (distance from the origin to triangle vertices) at height $z$:

$$
r(z) = r_t + (l - r_t) \cdot \frac{z}{h}.
$$

A point in the triangle at height $z$ is generated using barycentric coordinates $(u, v)$ to ensure uniform sampling within the triangular cross-section:

$$
u, v \sim \mathcal{U}(0, 1), \quad \text{if } u + v > 1: u \leftarrow 1 - u,\ v \leftarrow 1 - v.
$$

The first three coordinates (triangle plane) are then:

$$
x_1 = r(z)(1 - u - v), \quad
x_2 = r(z) \cdot u, \quad
x_3 = r(z) \cdot v
$$

For dimensions $j = 4, \dots, p-1$, values are tapered linearly toward zero near the tip:

$$
x_j \sim \mathcal{U}(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right).
$$

The `gen_pyrtri(n, p, h, l, rt)` function generates a dataset of $n$ points forming a high-dimensional pyramid-like structure with a triangular cross-section. The structure is embedded in a $p$-dimensional space, with the tip located at height 0 and the base at height $h$. The triangle expands linearly in size from tip to base, with more points concentrated near the tip.

Let $x\_1, x\_2, \dots, x\_p$ denote the coordinates of the generated points. The final coordinate $x\_p$ encodes the height $z$ of each point and is drawn from an exponential distribution capped at $h$:

$$
x_p = z \sim \min\left(\text{Exp}(\lambda = 2/h),\ h\right).
$$

Let $r(z)$ denote the scaling factor (distance from the origin to triangle vertices) at height $z$:

$$
r(z) = r_t + (l - r_t) \cdot \frac{z}{h}.
$$

A point in the triangle at height $z$ is generated using barycentric coordinates $(u, v)$ to ensure uniform sampling within the triangular cross-section:

$$
u, v \sim \mathcal{U}(0, 1), \quad \text{if } u + v > 1: u \leftarrow 1 - u,\ v \leftarrow 1 - v.
$$

The first three coordinates (triangle plane) are then:

$$
x_1 = r(z)(1 - u - v), \quad
x_2 = r(z) \cdot u, \quad
x_3 = r(z) \cdot v
$$

For dimensions $j = 4, \dots, p-1$, values are tapered linearly toward zero near the tip:

$$
x_j \sim \mathcal{U}(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right).
$$

The result is a tapered, high-dimensional triangle-based pyramid with more points near the apex and greater width at the base.

@@

The `gen_pyrstar(n, p, h, rb)` function generates a dataset of \$n\$ points forming a high-dimensional pyramid-shaped structure with a hexagonal base. The pyramid extends vertically from height $z = 0$ (tip) to height $z = h$ (base), embedded in a $p$-dimensional space. The distribution concentrates more points near the base, and the hexagonal spread tapers toward the tip.

Let $x\_1, x\_2, \dots, x\_p$ denote the coordinates of the generated points. The final coordinate $x\_p$ encodes the height $z$:

$$
x_p = z \sim \mathcal{U}(0, h).
$$

The radius at height $z$ scales linearly from zero (tip) to the base radius $r\_b$:

$$
r(z) = r_b \cdot \left(1 - \frac{z}{h}\right).
$$

Each point is placed within a regular hexagon in the plane $(x\_1, x\_2)$, using a randomly chosen hexagon sector angle $\theta \in \left{0, \frac{\pi}{3}, \frac{2\pi}{3}, \pi, \frac{4\pi}{3}, \frac{5\pi}{3} \right}$ and a uniformly random radial scaling factor:

$$
\theta \sim \text{Uniform sample from 6 hexagon angles}, \quad
r_{\text{point}} \sim \sqrt{\mathcal{U}(0, 1)}.
$$

Then, the first two coordinates are:

$$
x_1 = r(z) \cdot r_{\text{point}} \cdot \cos(\theta), \quad
x_2 = r(z) \cdot r_{\text{point}} \cdot \sin(\theta)
$$

For dimensions $j = 3, \dots, p-1$, values are tapered toward zero near the tip:

$$
x_j \sim \mathcal{U}(-0.1, 0.1) \cdot \left(1 - \frac{z}{h} \right).
$$

This results in a smooth, star-like pyramid in $p$-dimensional space, with a broad hexagonal base and narrow tip.

@@

The `gen_pyrholes(n, p)` function generates $n$ points embedded in a $p$-dimensional simplex using a chaotic attractor-like midpoint algorithm. The result is a fractal-like structure that reveals holes or gaps in the data cloud, forming a "Sierpinski-like pyramid" in high dimensions.

Let $x\_1, x\_2, \dots, x\_p$ denote the coordinates of the generated points. The generation process begins with an initial point $T\_0 \in \[0, 1]^p$ drawn from a uniform distribution:

$$
T_0 \sim \mathcal{U}(0, 1)^p.
$$

Let $C\_1, C\_2, \dots, C\_{p+1}$ denote the corner vertices of a $p$-dimensional simplex. At each iteration $i = 1, \dots, n$, a new point is computed by taking the midpoint between the previous point $T\_{i-1}$ and a randomly selected vertex $C\_k$:

$$
T_i = \frac{1}{2}(T_{i-1} + C_k), \quad C_k \in \{C_1, \dots, C_{p+1}\}.
$$

This recursive midpoint rule generates self-similar patterns with systematic voids (holes) between clusters of points. The points remain bounded inside the convex hull of the simplex.

The final output is a $n \times p$ matrix where each row represents a point:

$$
X = \{T_1, T_2, \dots, T_n\}, \quad X \in \mathbb{R}^{n \times p}.
$$

@@

- Pyrholes: This structure is useful for testing clustering and NLDR algorithms on non-convex and sparse high-dimensional shapes.
